---
title: "Linear Models versus Linear Mixed Models"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This tutorial introduces the difference between ordinary linear models and linear mixed models using repeated-measures data. The aim is to understand when the independence assumption of the linear model is not appropriate, and how mixed models account for subject-level variation.

## Learning objectives

By the end of this tutorial, you should be able to:

- describe the assumptions of an ordinary linear model
- explain why repeated measures violate independence
- fit a linear model and a linear mixed model in R
- interpret fixed and random effects
- visualise the difference between population-level and subject-level fits

## Required packages

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom)
library(broom.mixed)
library(ggeffects)
library(emmeans)
library(performance)
library(patchwork)
library(cowplot)
library(modelsummary)
source("scripts/simulate_lmm_data.R")
```

## Mathematical background

### Ordinary linear model

For independent observations, the linear model can be written as

$$
y_i = \beta_0 + \beta_1 x_i + \varepsilon_i
$$

where

$$
\varepsilon_i \sim N(0, \sigma^2)
$$

The key assumption is that the residuals are independent.

In matrix notation, this becomes

$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\varepsilon}
$$

with

$$
\boldsymbol{\varepsilon} \sim N(\mathbf{0}, \sigma^2 \mathbf{I})
$$

Therefore,

$$
\mathrm{Var}(\mathbf{y}) = \sigma^2 \mathbf{I}
$$

This means that all observations have the same variance and are mutually uncorrelated.

### Why this becomes problematic for repeated measures

Suppose subject $j$ is measured several times, with observation $i$ indexing repeated measurements within subject. Then two observations from the same subject are usually more similar to each other than two observations from different subjects.

If we ignore that dependence and fit

$$
y_{ij} = \beta_0 + \beta_1 x_{ij} + \varepsilon_{ij}
$$

as though all observations were independent, standard errors and inferential results may be misleading.

### Linear mixed model with a random intercept

A random-intercept mixed model is

$$
y_{ij} = \beta_0 + \beta_1 x_{ij} + b_{0j} + \varepsilon_{ij}
$$

where:

- $\beta_0$ is the population-average intercept
- $\beta_1$ is the population-average slope
- $b_{0j}$ is the random intercept for subject $j$
- $\varepsilon_{ij}$ is the residual error

Assume

$$
b_{0j} \sim N(0, \tau_0^2)
\qquad \text{and} \qquad
\varepsilon_{ij} \sim N(0, \sigma^2)
$$

This allows each subject to have their own baseline level while sharing a common slope.

### Linear mixed model with a random intercept and random slope

If subjects also differ in how they change with $x$, then we can write

$$
y_{ij} = \beta_0 + \beta_1 x_{ij} + b_{0j} + b_{1j}x_{ij} + \varepsilon_{ij}
$$

where $b_{1j}$ is a subject-specific random slope.

We assume

$$
\begin{pmatrix}
b_{0j} \\
b_{1j}
\end{pmatrix}
\sim
N
\left(
\begin{pmatrix}
0 \\
0
\end{pmatrix},
\mathbf{G}
\right)
$$

with

$$
\mathbf{G} =
\begin{pmatrix}
\tau_0^2 & \tau_{01} \\
\tau_{01} & \tau_1^2
\end{pmatrix}
$$

This means that each subject can have both their own baseline and their own trajectory.

### Matrix form of the mixed model

The linear mixed model can be written as

$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{Z}\mathbf{b} + \boldsymbol{\varepsilon}
$$

where:

- $\mathbf{X}\boldsymbol{\beta}$ is the fixed-effects component
- $\mathbf{Z}\mathbf{b}$ is the random-effects component

Assume

$$
\mathbf{b} \sim N(\mathbf{0}, \mathbf{G})
\qquad \text{and} \qquad
\boldsymbol{\varepsilon} \sim N(\mathbf{0}, \mathbf{R})
$$

Then

$$
\mathrm{Var}(\mathbf{y}) = \mathbf{Z}\mathbf{G}\mathbf{Z}^{\top} + \mathbf{R}
$$

This is the key difference from the ordinary linear model, where

$$
\mathrm{Var}(\mathbf{y}) = \sigma^2 \mathbf{I}
$$

In other words, the mixed model explicitly accounts for covariance induced by clustering or repeated measurements.

## Simulate repeated-measures data

```{r}
set.seed(123)

n_subjects <- 30
n_time <- 6

subject_df <- tibble(
  subject = factor(1:n_subjects),
  b0 = rnorm(n_subjects, mean = 0, sd = 6),
  b1 = rnorm(n_subjects, mean = 0, sd = 1.2)
)

df <- expand_grid(
  subject = factor(1:n_subjects),
  time = 0:(n_time - 1)
) |>
  left_join(subject_df, by = "subject") |>
  mutate(
    y = 20 + 2.5 * time + b0 + b1 * time + rnorm(n(), mean = 0, sd = 2)
  )

glimpse(df)
```

## Visualise the raw data

```{r}
p_raw <- ggplot(df, aes(x = time, y = y, colour = subject)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_line(alpha = 0.5) +
  guides(colour = "none") +
  labs(
    title = "Repeated-measures data",
    subtitle = "Each colour represents one subject",
    x = "Time",
    y = "Outcome"
  ) +
  theme_minimal(base_size = 13)

p_raw
```

The plot already suggests that observations from the same subject are not independent: each subject tends to follow their own trajectory.

## Fit an ordinary linear model

```{r}
m_lm <- lm(y ~ time, data = df)
summary(m_lm)
```

This model assumes that all observations are independent and ignores the subject structure entirely.

## Fit a linear mixed model with a random intercept

```{r}
m_lmm_int <- lmer(y ~ time + (1 | subject), data = df)
summary(m_lmm_int)
```

This model allows each subject to have their own baseline value, while keeping a common slope for time.

## Fit a linear mixed model with a random intercept and slope

```{r}
m_lmm_slope <- lmer(y ~ time + (time | subject), data = df)
summary(m_lmm_slope)
```

This model allows each subject to have both their own baseline and their own time trend.

## Compare the fitted values

```{r}
df <- df |>
  mutate(
    pred_lm = predict(m_lm),
    pred_lmm_int = predict(m_lmm_int),
    pred_lmm_slope = predict(m_lmm_slope)
  )
```

### Linear model fit

```{r}
p_lm <- ggplot(df, aes(time, y, colour = subject)) +
  geom_point(alpha = 0.5) +
  geom_abline(
    intercept = coef(m_lm)[1],
    slope = coef(m_lm)[2],
    linewidth = 1.1,
    colour = "black"
  ) +
  guides(colour = "none") +
  labs(
    title = "Ordinary linear model",
    subtitle = "One global line for all observations",
    x = "Time",
    y = "Outcome"
  ) +
  theme_minimal(base_size = 13)

p_lm
```

### Random-intercept model fit

```{r}
p_lmm_int <- ggplot(df, aes(time, y, colour = subject)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = pred_lmm_int, group = subject), linewidth = 0.9) +
  guides(colour = "none") +
  labs(
    title = "Mixed model with random intercept",
    subtitle = "Different subject baselines, common slope",
    x = "Time",
    y = "Outcome"
  ) +
  theme_minimal(base_size = 13)

p_lmm_int
```

### Random-intercept and random-slope model fit

```{r}
p_lmm_slope <- ggplot(df, aes(time, y, colour = subject)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = pred_lmm_slope, group = subject), linewidth = 0.9) +
  guides(colour = "none") +
  labs(
    title = "Mixed model with random intercept and slope",
    subtitle = "Different subject baselines and slopes",
    x = "Time",
    y = "Outcome"
  ) +
  theme_minimal(base_size = 13)

p_lmm_slope
```

### Side-by-side comparison

```{r}
cowplot::plot_grid(
  p_lm,
  p_lmm_int,
  p_lmm_slope,
  labels = c("A", "B", "C"),
  nrow = 1,
  align = "hv"
)
```

This comparison is central:

- the ordinary linear model fits a single global line
- the random-intercept model allows subject-specific offsets
- the random-slope model allows subject-specific trajectories

## Population-level predictions

```{r}
p_pred <- ggpredict(m_lmm_slope, terms = "time") |>
  plot() +
  labs(
    title = "Population-level predicted relationship",
    x = "Time",
    y = "Predicted outcome"
  ) +
  theme_minimal(base_size = 13)

p_pred
```

## Compare model summaries

```{r}
modelsummary(
  list(
    "LM" = m_lm,
    "LMM: random intercept" = m_lmm_int,
    "LMM: random intercept + slope" = m_lmm_slope
  ),
  statistic = "({std.error})"
)
```

## Diagnostics

```{r}
#| fig-width: 14
#| fig-height: 14
#| dpi: 300
#| out-width: "100%"

cm <- performance::check_model(m_lmm_slope)
plot(cm, n_columns = 2)
```

## Interpretation

### Fixed effects

The fixed effect of time represents the average relationship between time and outcome across the population.

### Random effects

The random effects describe how subjects vary around the population-average relationship.

- random intercepts capture differences in baseline levels
- random slopes capture differences in time trajectories

### Key conceptual difference

In an ordinary linear model, unexplained between-subject variability is absorbed into the residual term.

In a mixed model, part of that variability is modelled explicitly through random effects.

This is the main reason mixed models are appropriate for clustered or repeated-measures data.

## Take-home message

Use an ordinary linear model when observations are independent.

Use a linear mixed model when observations are grouped or repeated within higher-level units such as subjects, classes, sites, or batches.

The mixed model extends the linear model by explicitly modelling the covariance structure induced by clustering.