---
title: "Outlier identification and treatment in R with `performance` and capping"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This document shows a practical workflow for outlier identification in R using the `performance` package, followed by treatment options with a focus on capping. The workflow is designed for continuous variables.

The recommended sequence is:

1. identify candidate outliers using `performance::check_outliers()`
2. review which rows and variables are affected
3. create a capped version of selected variables
4. compare descriptive summaries before and after capping
5. refit the model on the treated dataset as a sensitivity analysis

## Libraries

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(performance)
  library(lme4)
})
```

## Example data

Replace this section with your own dataset if needed.

```{r}
set.seed(123)

n_subjects <- 300
n_time <- 6

subject_df <- tibble(
  subject = factor(seq_len(n_subjects)),
  age = rnorm(n_subjects, mean = 55, sd = 10),
  bmi = rnorm(n_subjects, mean = 27, sd = 4),
  b0 = rnorm(n_subjects, mean = 0, sd = 5),
  b1 = rnorm(n_subjects, mean = 0, sd = 1.0)
)

df <- expand_grid(
  subject = factor(seq_len(n_subjects)),
  time = 0:(n_time - 1)
) |>
  left_join(subject_df, by = "subject") |>
  mutate(
    biomarker = 50 + 1.8 * time + 0.20 * age + 0.35 * bmi + b0 + b1 * time + rnorm(n(), 0, 2)
  )

# add a few artificial extreme values for demonstration
set.seed(999)
idx_biomarker <- sample(seq_len(nrow(df)), size = 12)
df$biomarker[idx_biomarker] <- df$biomarker[idx_biomarker] + rnorm(length(idx_biomarker), mean = 35, sd = 8)

idx_bmi <- sample(seq_len(nrow(df)), size = 8)
df$bmi[idx_bmi] <- df$bmi[idx_bmi] + 18

glimpse(df)
```

## Variables to screen for outliers

Outlier detection should usually be limited to continuous variables that are scientifically plausible candidates for extreme values.

```{r}
vars_to_check <- c("biomarker", "age", "bmi", "time")
```

## Identify outliers with `performance::check_outliers()`

This version uses only the robust z-score method, which avoids extra package dependencies.

```{r}
outlier_result <- check_outliers(
  df[, vars_to_check],
  method = "zscore_robust"
)

outlier_result
```

## Optional plot of the outlier result

If the `see` package is installed, `plot()` can visualise the outlier result.

```{r}
if (requireNamespace("see", quietly = TRUE)) {
  plot(outlier_result)
}
```

## Extract row level outlier information

```{r}
outlier_df <- as.data.frame(outlier_result)

head(outlier_df)
```

## Append outlier flags back to the dataset

For a single method, `Outlier` is typically a logical flag or a 0/1 style indicator.

```{r}
df_flagged <- df |>
  mutate(
    row_id = seq_len(n()),
    outlier_flag_raw = outlier_df$Outlier,
    is_outlier = as.logical(outlier_flag_raw)
  )

df_flagged |>
  count(is_outlier)
```

## Inspect candidate outliers

```{r}
candidate_outliers <- df_flagged |>
  filter(is_outlier) |>
  select(row_id, subject, time, all_of(vars_to_check)) |>
  arrange(subject, time)

candidate_outliers
```

## Simple visual checks

### Histogram of the main variable

```{r}
ggplot(df_flagged, aes(x = biomarker)) +
  geom_histogram(bins = 40) +
  labs(
    title = "Distribution of biomarker before treatment",
    x = "Biomarker",
    y = "Count"
  )
```

### Boxplot with flagged rows highlighted

```{r}
ggplot(df_flagged, aes(x = "", y = biomarker, colour = is_outlier)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.12, alpha = 0.5) +
  labs(
    title = "Biomarker values with outlier flags",
    x = NULL,
    y = "Biomarker"
  )
```

### Scatter plot over time with flagged rows highlighted

```{r}
ggplot(df_flagged, aes(x = time, y = biomarker, colour = is_outlier)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Biomarker over time with flagged rows highlighted",
    x = "Time",
    y = "Biomarker"
  )
```

## Capping function

This helper applies percentile-based capping, also called winsorisation. Values below the lower percentile are set to the lower bound, and values above the upper percentile are set to the upper bound.

```{r}
cap_percentile <- function(x, probs = c(0.01, 0.99), na.rm = TRUE) {
  stopifnot(is.numeric(x), length(probs) == 2)
  qs <- quantile(x, probs = probs, na.rm = na.rm, names = FALSE)
  pmin(pmax(x, qs[1]), qs[2])
}
```

## Apply capping to selected variables

Here the example caps `biomarker` and `bmi` at the 1st and 99th percentiles.

```{r}
df_capped <- df_flagged |>
  mutate(
    biomarker_capped = cap_percentile(biomarker, probs = c(0.01, 0.99)),
    bmi_capped = cap_percentile(bmi, probs = c(0.01, 0.99))
  )
```

## Compare before and after capping

```{r}
comparison_tab <- tibble(
  variable = c("biomarker", "biomarker_capped", "bmi", "bmi_capped"),
  min = c(
    min(df_capped$biomarker, na.rm = TRUE),
    min(df_capped$biomarker_capped, na.rm = TRUE),
    min(df_capped$bmi, na.rm = TRUE),
    min(df_capped$bmi_capped, na.rm = TRUE)
  ),
  p01 = c(
    quantile(df_capped$biomarker, 0.01, na.rm = TRUE),
    quantile(df_capped$biomarker_capped, 0.01, na.rm = TRUE),
    quantile(df_capped$bmi, 0.01, na.rm = TRUE),
    quantile(df_capped$bmi_capped, 0.01, na.rm = TRUE)
  ),
  median = c(
    median(df_capped$biomarker, na.rm = TRUE),
    median(df_capped$biomarker_capped, na.rm = TRUE),
    median(df_capped$bmi, na.rm = TRUE),
    median(df_capped$bmi_capped, na.rm = TRUE)
  ),
  p99 = c(
    quantile(df_capped$biomarker, 0.99, na.rm = TRUE),
    quantile(df_capped$biomarker_capped, 0.99, na.rm = TRUE),
    quantile(df_capped$bmi, 0.99, na.rm = TRUE),
    quantile(df_capped$bmi_capped, 0.99, na.rm = TRUE)
  ),
  max = c(
    max(df_capped$biomarker, na.rm = TRUE),
    max(df_capped$biomarker_capped, na.rm = TRUE),
    max(df_capped$bmi, na.rm = TRUE),
    max(df_capped$bmi_capped, na.rm = TRUE)
  )
)

comparison_tab
```

## Plot before and after capping

### Biomarker

```{r}
plot_bio <- df_capped |>
  select(biomarker, biomarker_capped) |>
  pivot_longer(everything(), names_to = "version", values_to = "value")

ggplot(plot_bio, aes(x = value)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ version, scales = "free_y") +
  labs(
    title = "Biomarker before and after capping",
    x = "Value",
    y = "Count"
  )
```

### BMI

```{r}
plot_bmi <- df_capped |>
  select(bmi, bmi_capped) |>
  pivot_longer(everything(), names_to = "version", values_to = "value")

ggplot(plot_bmi, aes(x = value)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ version, scales = "free_y") +
  labs(
    title = "BMI before and after capping",
    x = "Value",
    y = "Count"
  )
```

## Optional model sensitivity analysis

A good practice is to compare the substantive model fitted to the original and capped datasets.

```{r}
fit_original <- lmer(
  biomarker ~ time + age + bmi + (time | subject),
  data = df_capped,
  REML = FALSE
)

fit_capped <- lmer(
  biomarker_capped ~ time + age + bmi_capped + (time | subject),
  data = df_capped,
  REML = FALSE
)

coef_tab <- tibble(
  term = names(fixef(fit_original)),
  estimate_original = unname(fixef(fit_original)),
  estimate_capped = unname(fixef(fit_capped)),
  delta = estimate_capped - estimate_original
)

coef_tab
```

## Alternative treatment options

Capping is only one treatment strategy. Depending on the analysis plan and subject matter context, alternatives include:

- leaving values unchanged but documenting influence
- transforming variables, for example using a log transform
- setting implausible values to missing and using an appropriate missing-data strategy
- removing clearly erroneous records after documented review

## Minimal adaptation to your own data

For your own dataset, you would usually only need to change:

```{r eval=FALSE}
vars_to_check <- c("outcome", "covariate1", "covariate2")

outlier_result <- check_outliers(
  mydata[, vars_to_check],
  method = "zscore_robust"
)

mydata2 <- mydata |>
  mutate(
    outcome_capped = cap_percentile(outcome, probs = c(0.01, 0.99)),
    covariate1_capped = cap_percentile(covariate1, probs = c(0.01, 0.99))
  )
```

## Reporting text

You can use the following wording in a methods section:

Potential outliers were screened using `performance::check_outliers()` with the robust z-score method. As a sensitivity analysis, selected continuous variables were capped at the 1st and 99th percentiles, and the substantive model was refitted using the capped variables.